BOLD Parameter Estimation Using Particle Filters

### 2010-07-05 doc/thesis/methods.tex
some mild additions to methods and results
more work....

\section{Resampling}
The algorithm for resampling is described in \autoref{sec:Particle Filter Resampling}.
The primary design decision for resampling is the regularizing
kernel. As mentioned previously, the Gaussian kernel is extremely convenient,
because it is very simple to sample from. As discussed in \autoref{sec:Particle Filter Resampling},
as long as resampling is kept as a last resort, some amount of over-smoothing
won't impair convergence. Therefore, for this work I chose a Gaussian kernel of
bandwidth equal to the original distribution's covariance. Obviously this will
apply a rather large amount of smoothing to the distribution; however, on average
resampling is only applied every 20 to 30 measurements, and because randomization
is being applied to model updates this gives the filter some mobility.

@@

\section{Simulation}
I performed two types of simulations. First, I simulated a single BOLD time-series based
on a randomly selected set of model parameters. This process was relatively straight forward,
given the state-space equations for the BOLD signal. After a "true" signal was generated,
I then added a carrier level, since BOLD is typically measured as a \% difference from the
base level. I tested the particle filter with two types of noise; independent and identically
distributed gaussian noise and weiner drift. For both I tested with various standard deviations.

Once this noisy simulated time series was generated, the exact same particle filter algorithm
that would later be run on full sized images, was run on this single voxel image. I ran
a series of tests to determine the convergence rate of the particle filter, the number
of particles that were required, how weighting functions compared, how different de-trending
methods compared with each other and, finally the variance of the result. By running the exact
time-series with different noise realizations, it was possible to determine the model variance.
As the reader may know, the error of an estimator may be calculated as:
\begin{equation}
MSE(\Theta) = Var(\Theta) + Bias(\Theta)^2
\end{equation}
The variance is an expression of how much the result would change for different noise realizations,
whereas the bias is an expression of how well the model matches the true underlying model. In
this case, because the same model is being used in the particle filter and underlying simulation,
the bias is actually zero. Obviously when this is calculated using \emph{real} data with an unknown
underlying state space equation, there will be some amount of bias error, but assuming that the
noise is similar to the noise used in these tests, the model variance will actually be about the
same. Thus calculating the model variance is extremely helpful in calculating how well determined
our model is, and how consistent it will be for real data. A single timeseries, as opposed to the
thousands present in a real image, makes it easier to 
compare the output with the ground truth, with various parameters.

Second I used a modified version of the FSL tool 
POSSUM to generate an entire FMRI image from a parameter map. The parameter map was generated
by taking an existing activation map and assigning discrete parameter sets to each region.
The result was a four dimensional (length x width
x height x parameter) image with spatially varying parameters. Possum was then modified
to take a parameter map and generate activation levels depending on the parameters at that
point. The patch for POSSUM will be made available. As an unfortunate side effect of 
not using Possums' original activation scale, I manually added 750 to the total level of
simulated Possum images. This is because the BOLD \% difference levels were in the range
of 50 - 100\% from the base, about 5 times as large as they should have been. Ultimately
this should not have an effect on the parameters (other than perhaps $\epsilon$ and $V_0$). 
For each time-series in the simulated FMRI image, the final \emph{static} parameters were saved
into a parameter map. This parameter map may then be compared to the map used to generate the 
simulated data; additionally a new simulation using the calculated parameters may also be 
generated to test the difference in activation levels between the real parameters and the
estimated ones. Since it is clear that the parameters are not fully orthogonal 
(\cite{Deneux2006}), its possible that two sets of parameters are functionally equivalent,
but have different parameters. This way, an absolute 
quantitative difference between the two parameter sets may be found.

\section{Real Data}
Finally, we also performed inference based on real FMRI data. The scanner we used...
%... more specifics...

Before performing tests on a full image, I tested the results of the particle filter
on regions deemed active and non-active by statistical parametric mapping.

>>>>

\section{Resampling}
\label{sec:Resampling}
The algorithm for resampling is described in \autoref{sec:Particle Filter Resampling}.
The primary design decision for resampling is the regularizing
kernel. As mentioned previously, the Gaussian kernel is extremely convenient,
because it is very simple to sample from. As discussed in \autoref{sec:Particle Filter Resampling},
as long as resampling is kept as a last resort, some amount of over-smoothing
won't impair convergence. Therefore, for this work I chose a Gaussian kernel of
bandwidth equal to the original distribution's covariance. Obviously this will
apply a rather large amount of smoothing to the distribution; however, on average
resampling is only applied every 20 to 30 measurements, and because randomization
is being applied to model updates this gives the filter some mobility.

@@

### 2010-07-05 doc/thesis/particlefilters.tex
almost done with methods
more work....

\section{Weighting Function}
Because the distribution of $\nu_y$ in \autoref{eq:measass} is unknown,
it is necessary to choose a distribution for this. This distribution,
is extremely important because $\nu_y \sim P(y_k | x(T))$, which is used
for updating weights. Ideally this weighting function would exactly 
match the measurement error in
the output. It is logical to assume this error is centered at zero and 
has a scale comparable to the signal levels.
While a Gaussian function is the traditional choice, I wanted to try other
distributions as well, given the unpredictable nature of the noise present in FMRI.
The choice of this function will be discussed further in \autoref{sec:Methods
Weighting Function}.

>>>>

\section{Weighting Function}
Because the distribution of $\nu_y$ in \autoref{eq:measass} is unknown,
it is necessary to choose a distribution for this. This distribution
is extremely important because $\nu_y \sim P(y_k | x(T))$, which is used
for updating weights. Ideally this weighting function would exactly 
match the measurement error in
the output. It is logical to assume this error is centered at zero and 
has a scale comparable to the signal levels.
While a Gaussian function is the traditional choice, I wanted to try other
distributions as well, given the unpredictable nature of the noise present in FMRI.
The choice of this function will be discussed further in \autoref{sec:Methods
Weighting Function}.

\begin{figure}
\includegraphics[width=16cm]{images/particle_filter}
\caption{Particle Filter progression, note that the initial support is flat; the particles
are equally spaced between -10 and 10}
\end{figure}

### 2010-07-05 doc/thesis/results.tex
more work....

\chapter{Results}
\section{Single-Voxel Simulation}
The results

>>>>

\chapter{Results}

\section{Simulation}
I performed two types of simulations. First, I simulated a single BOLD time-series based
on a randomly selected set of model parameters. This process was relatively straight forward,
given the state-space equations for the BOLD signal. After a "true" signal was generated,
I then added a carrier level, since BOLD is typically measured as a \% difference from the
base level. I tested the particle filter with two types of noise; independent and identically
distributed gaussian noise and weiner drift. For both I tested with various standard deviations.

Once this noisy simulated time series was generated, the exact same particle filter algorithm
that would later be run on full sized images, was run on this single voxel image. I ran
a series of tests to determine the convergence rate of the particle filter, the number
of particles that were required, how weighting functions compared, how different de-trending
methods compared with each other and, finally the variance of the result. By running the exact
time-series with different noise realizations, it was possible to determine the model variance.
As the reader may know, the error of an estimator may be calculated as:
\begin{equation}
MSE(\Theta) = Var(\Theta) + Bias(\Theta)^2
\end{equation}
The variance is an expression of how much the result would change for different noise realizations,
whereas the bias is an expression of how well the model matches the true underlying model. In
this case, because the same model is being used in the particle filter and underlying simulation,
the bias is actually zero. Obviously when this is calculated using \emph{real} data with an unknown
underlying state space equation, there will be some amount of bias error, but assuming that the
noise is similar to the noise used in these tests, the model variance will actually be about the
same. Thus calculating the model variance is extremely helpful in calculating how well determined
our model is, and how consistent it will be for real data. A single timeseries, as opposed to the
thousands present in a real image, makes it easier to 
compare the output with the ground truth, with various parameters.

Second I used a modified version of the FSL tool 
POSSUM to generate an entire FMRI image from a parameter map. The parameter map was generated
by taking an existing activation map and assigning discrete parameter sets to each region.
The result was a four dimensional (length x width
x height x parameter) image with spatially varying parameters. Possum was then modified
to take a parameter map and generate activation levels depending on the parameters at that
point. The patch for POSSUM will be made available. As an unfortunate side effect of 
not using Possums' original activation scale, I manually added 750 to the total level of
simulated Possum images. This is because the BOLD \% difference levels were in the range
of 50 - 100\% from the base, about 5 times as large as they should have been. Ultimately
this should not have an effect on the parameters (other than perhaps $\epsilon$ and $V_0$). 
For each time-series in the simulated FMRI image, the final \emph{static} parameters were saved
into a parameter map. This parameter map may then be compared to the map used to generate the 
simulated data; additionally a new simulation using the calculated parameters may also be 
generated to test the difference in activation levels between the real parameters and the
estimated ones. Since it is clear that the parameters are not fully orthogonal 
(\cite{Deneux2006}), its possible that two sets of parameters are functionally equivalent,
but have different parameters. This way, an absolute 
quantitative difference between the two parameter sets may be found.

\section{Real Data}
Finally, we also performed inference based on real FMRI data. The scanner we used...
%... more specifics...

Before performing tests on a full image, I tested the results of the particle filter
on regions deemed active and non-active by statistical parametric mapping. 
\section{Single-Voxel Simulation}
The results
